Estimation of Spatiotemporal Integro-Difference Equations for Dynamical Neural Field Models

### 2009-11-11 ltx/BrainIDE.tex
Added references about neural models in intro of tex file andded refs to bib file

\IEEEPARstart{T}{he} human brain is arguably the world's most complex system. Approximately 100 billion neurons and 60 trillion synapses operate in concert to process information, resulting in the cognition that determines our behavior. The brain's efficiency, robustness, and adaptability are unparalleled by any man-made device and, despite decades of concerted research, our understanding of its complex dynamics remains modest. This has led researchers to study different scales of dynamics: starting from individual proteins, synapses, and neurons; moving through to neuronal networks and ensembles of neuronal networks. Whilst our understanding of the function of neurons is well developed, the overall behavior of the brain's meso and macro-scale dynamics remains largely a mystery.  Understanding the brain at this level is extremely important since this is the level where pathologies such as epilepsy, Parkinson's disease and schizophrenia are manifested.

To date, there has been a considerable volume of work in generating physiologically plausible neural field models to fill the void of understanding brain dynamics at this meso/macroscopic scale. Mathematical neural field  models have provided insights into the underlying physics and dynamics of electroencephalography (EEG) and magnetoencephalography (MEG) (see~\cite{Deco2008}~\cite{David2003} for recent reviews). These models have demonstrated possible mechanisms for the genesis of neural rhythms (such as the alpha and gamma rhythms) \cite{Liley1999} \cite{RENNIE2000}, epileptic seizure generation \cite{DaSilva2003},~\cite{Suffczynski2004} and~\cite{Wendling2005} and insights into other pathologies (ref, Parkinsons Scitzo) that would be impossible to gain from experimental data . Unfortunately, to date the use of these models in the clinic has very limited, since these models are very general and pathologies almost always have unique underlying patient specific causes. Data from EEG and functional magnetic resonance imaging (fMRI) offers the patient specific link to macro-scale cortical dynamics, making these tools readily applicable to the clinic. However the underlying system properties, or system states, are hidden from EEG and fMRI data, making predictions of the underlying physiology inherently difficult.

For models to be clinically viable they must be patient specific. A possible approach to achieve this would be to use a general neural field model, like the \textcolor[rgb]{1.00,0.00,0.00}{Wilson Cohen Jansen and Ritt model ref}, and fit it to a patients EEG data. Fitting the general neural field models to individuals is a highly non-trivial task, and until very recently this has not been reported in the literature. This is most likely due to the large parameter set that most models use in order to be physiologically plausible. To our knowledge, the first (and only) published research that attempts parameter estimation with a neural field model used an unscented Kalman filter to track parameters of the Wilson-Cowan neural field equations~\cite{schiff2008kalman}.  This work marks the first step in what has the potential to revolutionize the treatment of many neurological diseases where therapeutic electrical stimulation is viable. Generating patient specific models allows that application of a range of techniques from control systems theory, where tailored electrical stimulation could be used therapeutically in a closed loop fashion.  Currently available epileptic seizure prediction and control devices (i.e., the vagal nerve stimulator) are implemented in an ``open loop".  That is, the therapeutic electrical stimulation waveforms are adjusted for each patient by trial and error, disregarding the patient's neurodynamics and information about their particular pathologies. Given access to an accurate model, the application of control theory in these circumstances would allow for robust therapeutic stimulations.
The work from Schiff and Sauer~\cite{schiff2008kalman} successfully demonstrated  estimation of two parameters from the WC equations, whilst the other parameters were fixed. This motivates the question, what parameters are the most important? How patient specific do the assumed parameters have to be for the model to be useful? In this paper we address both of these questions and provide a theoretical platform to perform patient specific grey-box modeling of the human neocortex.

@@

The neural field models use statistics of the cortex to relate mean firing rates of pre-synaptic neural populations to post-synaptic mean membrane potentials. Each neural population represents a functional cortical processing unit referred to as a column. The columnar organization of the cortex is not discrete, but is continuous, where pyramidal cells are members of many columns. In general, cortical structure can be modeled in a physiologically plausible manner as being locally homogeneous (in short range intracortical connectivity) and heterogeneous (in long range corticocortical and corticothalamic connectivity)~\cite{Jirsa2009}~\cite{Qubbaj2007}. Locally, each column is thought to have symmetric short range local excitation, with surround inhibition~\cite{Braitenberg1998}. For example, this structural organisation is most studied in the visual system, where the surrounding inhibition effectively tunes column to a particular receptive visual field~\cite{Sullivan2006}.
Recent studies using neural field models have demonstrated the theoretical implications of the connectivity structure, where the connectivity kernel governs the bifurcation points of the models~\cite{Hutt2005} and types of oscillations that can be generated~\cite{Schmidt2009}. This implies that if we could estimate the connectivity structure for individuals we may be able to model patient specific dynamics.

Estimating function cortical connectivity via EEG measures is currently a highly active area of research. There is a lot of interest in understanding the hierarchy of brain regions are involved in specific tasks. Most works are based on auto-regressive (AR) modeling of the times series. Studies have concentrated on finding functional connectivity patterns using information contained in the AR coefficients. This approach has yielded interesting results but is limited in the sense that structure is not estimated hence this work cannot be related to the theoretical bifurcation analysis and the underlying system structure. (Granger causality, DTF, partial coherence)
An alternative approach to estimating the connectivity structure between brain regions is dynamic causal modeling (DCM). The DCM approach essentially studies the causality of information transform between pre-determined brain regions. DCM asks what connectivity structure between brain regions best explains a data set. The approach has generated several publications (ref) and generated evidence in understanding specific hierarchical networks of neural information processing.
Changes in connectivity would reflect learning.
Longrange connectivity can be estimated using DTI~\cite{Knock2009}, however this short range connectivity can not..
In this work we aim to estimate local intracortical connectivity. This is most invariant.
\section{Neural Field Model}

The Wilson-Cowan model has shown to be descriptive of a range of neurodynamics of the cortex. The model is capable of generating patterns of activity such as Turing patterns, spirals, .... The general form of the model is described here.
The post-synaptic potentials generated at a neuronal population $s$ by action potentials arriving from all other connected populations at position $r$ can be described by
\begin{equation}\label{Eq1}
v\left( {s,t} \right) = \int_{ - \infty }^t {h\left( {t - t'} \right)J\left( {r,t'} \right)dt'}.
\end{equation}
The post-synaptic response kernel $h(t)$ is described by
\begin{equation}\label{Eq2}
h(t) = \left\{ {\begin{array}{*{20}{c}}
   {\exp ( - {t \mathord{\left/
 {\vphantom {t \tau }} \right.
 \kern-\nulldelimiterspace} \tau })} & {t \ge 0}  \\
   0 & {t < 0}  \\
\end{array}} \right.,
\end{equation}
where $\tau$ is the synaptic time constant and typically has values from 10-20~ms (\textcolor[rgb]{1.00,0.00,0.00}{refs}). The synaptic response kernel can be more elaborate, using a rise and fall time constant, however this simple form serves our purposes. Equation~\ref{Eq1} relates average firing rate of a presynaptic neural population to the mean membrane potential of the post-synaptic neurons. The mean firing rate, $J$, is convolved with the response kernel to give membrane potential since the firing rate is assumed to follow a (homogeneous?) Poisson process. Nonlocal interactions between cortical populations are described by	
\begin{equation}\label{Eq3}
J\left( {s,t} \right) = \int_\Omega  {W\left( {s,r} \right)P\left( {v\left( {r,\bar t} \right)} \right)dr}  + E\left( {s,t} \right)
\end{equation}
where $\bar t = t - \frac{{\left| {s - r} \right|}}{d}$, $P$ is a mean pulse rate of population at $y$, $W$ is the connectivity kernel, $E$ is an external input, $\Omega$ is the spatial domain and $d$ is the propagation velocity from $r$ to $s$.

>>>>

\IEEEPARstart{T}{he} human brain is arguably the world's most complex system. Approximately 100 billion neurons and 60 trillion synapses operate in concert to process information, resulting in the cognition that determines our behavior. The brain's efficiency, robustness, and adaptability are unparalleled by any man-made device and, despite decades of concerted research, our understanding of its complex dynamics remains modest. This has led researchers to study different scales of dynamics: starting from individual proteins, synapses, and neurons; moving through to neuronal networks and ensembles of neuronal networks. Whilst our understanding of the function of neurons is well developed, the overall behavior of the brain's meso and macro-scale dynamics remains largely a mystery.  Understanding the brain at this level is extremely important since this is the scale where pathologies such as epilepsy, Parkinson's disease and schizophrenia are manifested.

To date, there has been a considerable volume of work in generating physiologically plausible neural field models to fill the void of understanding brain dynamics at the meso/macroscopic scale. Mathematical neural field  models have provided insights into the underlying physics and dynamics of electroencephalography (EEG) and magnetoencephalography (MEG) (see~\cite{Deco2008}~\cite{David2003} for recent reviews). These models have demonstrated possible mechanisms for the genesis of neural rhythms (such as the alpha and gamma rhythms) \cite{Liley1999} \cite{RENNIE2000}, epileptic seizure generation \cite{DaSilva2003},~\cite{Suffczynski2004} and~\cite{Wendling2005} and insights into other pathologies~\cite{Moran2008} that would be impossible to gain from experimental data alone. Unfortunately, to date the use of these models in the clinic has very limited, since the neural field model are constructed for a general brain and pathologies almost always have unique underlying patient specific causes. Data from EEG and functional magnetic resonance imaging (fMRI) offers the patient specific link to macro-scale cortical dynamics, making these tools readily applicable to the clinic. However the underlying system properties, or system states, are hidden from EEG and fMRI data, making predictions of the underlying physiology inherently difficult.

For models to be clinically viable they must be patient specific. A possible approach to achieve this would be to use a general neural field model, like the Wilson and Cohen~\cite{Wilson1973} or Jansen and Ritt model~\cite{Jansen1995}, and fit it to a patients EEG data. Fitting the general neural field models to individuals is a highly non-trivial task, and until very recently this has not been reported in the literature. This is most likely due to the large parameter set that most models use in order to be physiologically plausible. To our knowledge, the first (and only) published research that attempts parameter estimation with a neural field model used an unscented Kalman filter to track parameters of the Wilson-Cowan neural field equations~\cite{schiff2008kalman}.  This work marks the first step in what has the potential to revolutionize the treatment of many neurological diseases where therapeutic electrical stimulation is viable. Generating patient specific models allows that application of a range of techniques from control systems theory, where tailored electrical stimulation could be used therapeutically in a closed loop fashion.  Currently available epileptic seizure prediction and control devices (i.e., the vagal nerve stimulator) are implemented in an ``open loop".  That is, the therapeutic electrical stimulation waveforms are adjusted for each patient by trial and error, disregarding the patient's neurodynamics and information about their particular pathologies. Given access to an accurate model, the application of control theory in these circumstances would allow for robust therapeutic stimulations.
The work from Schiff and Sauer~\cite{schiff2008kalman} successfully demonstrated  estimation of two parameters from the WC equations, whilst the other parameters were fixed. This motivates the question, what parameters are the most important? How patient specific do the assumed parameters have to be for the model to be useful? In this paper we address both of these questions and provide a theoretical platform to perform patient specific grey-box modeling of the human neocortex.

@@

The neural field models use statistics of the cortex to relate mean firing rates of pre-synaptic neural populations to post-synaptic mean membrane potentials. Each neural population represents a functional cortical processing unit referred to as a column. The columnar organization of the cortex is not discrete, but is continuous, where pyramidal cells are members of many columns. In general, cortical structure can be modeled in a physiologically plausible manner as being locally homogeneous (in short range intracortical connectivity) and heterogeneous (in long range corticocortical and corticothalamic connectivity)~\cite{Jirsa2009}~\cite{Qubbaj2007}. Locally, each column is thought to have symmetric short range local excitation, with surround inhibition~\cite{Braitenberg1998}. For example, this structural organization is most studied in the visual system, where the surrounding inhibition effectively tunes column to a particular receptive visual field~\cite{Sullivan2006}.
Recent studies using neural field models have demonstrated the theoretical implications of the connectivity structure, where the connectivity kernel governs the bifurcation points of the models~\cite{Hutt2005} and types of oscillations that can be generated~\cite{Schmidt2009}. This implies that if we could estimate the connectivity structure for individuals we may be able to model patient specific dynamics.

Estimating function cortical connectivity via EEG measures is currently a highly active area of research. There is a lot of interest in understanding the hierarchy of brain regions are involved in specific tasks. Most works are based on auto-regressive (AR) modeling of the times series. Studies have concentrated on finding functional connectivity patterns using information contained in the AR coefficients(Granger causality, DTF, partial coherence). An alternative approach to estimating the connectivity structure between brain regions is dynamic causal modeling (DCM). The DCM approach essentially studies the causality of information transform between pre-determined brain regions. These regions are determined from prior information from fMRI, AR modeling or source inversion for example. DCM asks what connectivity structure between the specific isolated brain regions best explains a data set. The approach has generated several publications (ref) and generated evidence in understanding specific hierarchical networks of neural information processing. These approaches have yielded interesting results but is limited in the sense that structure is not estimated hence this work cannot be related to the theoretical bifurcation analysis and the underlying system structure.
Changes in connectivity would reflect learning.
Longrange connectivity can be estimated using DTI~\cite{Knock2009}, however this short range connectivity can not..
In this work we aim to estimate local intracortical connectivity. This is most invariant.

\section{Neural Field Model}
In this section we describe a variant of the Jansen and Ritt neural field model~\cite{Jansen1995} that will be used to create our connectivity estimator. This model has shown to be descriptive of a range of neurodynamics of the cortex such as evoked potentials and epileptic behavior and is capable of generating complex patterns of activity such as Turing patterns, spirals and traveling oscillations. The general form of the model is described here. The post-synaptic potentials generated at a neuronal population $s$ by action potentials arriving from all other connected populations at position $r$ can be described by
\begin{equation}\label{Eq1}
v\left( {s,t} \right) = \int_{ - \infty }^t {h\left( {t - t'} \right)J\left( {r,t'} \right)dt'}.
\end{equation}
The post-synaptic response kernel $h(t)$ is described by
\begin{equation}\label{Eq2}
h(t) = \left\{ {\begin{array}{*{20}{c}}
   {\exp ( - {t \mathord{\left/
 {\vphantom {t \tau }} \right.
 \kern-\nulldelimiterspace} \tau })} & {t \ge 0}  \\
   0 & {t < 0}  \\
\end{array}} \right.,
\end{equation}
where $\tau$ is the synaptic time constant and typically has values from 10-20~ms (\textcolor[rgb]{1.00,0.00,0.00}{refs}). The synaptic response kernel can be more elaborate, using a rise and fall time constant, however this simple form serves our purposes. Equation~\ref{Eq1} relates average firing rate of a presynaptic neural population to the mean membrane potential of the post-synaptic neurons. The mean firing rate, $J$, is convolved with the response kernel to give membrane potential since the firing rate is assumed to follow a (homogeneous?) Poisson process. Nonlocal interactions between cortical populations are described by	
\begin{equation}\label{Eq3}
J\left( {s,t} \right) = \int_\Omega  {W\left( {s,r} \right)P\left( {v\left( {r,\bar t} \right)} \right)dr}  + E\left( {s,t} \right)
\end{equation}
where $\bar t = t - \frac{{\left| {s - r} \right|}}{d}$, $P$ is a mean pulse rate of population at $y$, $W$ is the connectivity kernel, $E$ is an external input, $\Omega$ is the spatial domain and $d$ is the propagation velocity from $r$ to $s$.
